An Experimental Evaluation of Hybrid Vector Search (EA\&B)

### 2025-04-15 VLDB2026/main.tex
Update main.tex
add  peak memory DSG WIN UNIFY

	In recent years, hybrid query algorithms develop rapidly, giving rise to numerous attribute filtering ~\cite{NHQ,diskann} and range filtering algorithms \cite{serf,iRangeGraph}. In practical applications, the performance of hybrid query algorithms is influenced not only by unstructured data but also closely related to structured data. For attribute filtering algorithms, factors such as the number of base attributes, the number of query attributes, attribute distribution \cite{UNG}, and attribute selectivity \cite{ACORN} significantly impact algorithm performance. As for range filtering algorithms, the size of the query range and the characteristics of different datasets also play an important role in determining algorithm performance.

	Despite BigANN 2023 \cite{bigann2023} evaluates the performance of some algorithms, it presents several notable limitations:
	1) The evaluation covers a limited number of algorithms and fails to comprehensively include mainstream methods. For instance, attribute filtering algorithms such as NHQ and UNG are not included in the evaluation. 2) It focuses only on single-attribute and dual-attribute query scenarios, neglecting the more complex demands of multi-attribute queries. In real-world applications, an object typically involves multiple attributes. For example, in e-commerce scenarios, users often specify multiple attributes (e.g., color, brand, price range) simultaneously when searching.
	3) It does not evaluate algorithm performance under different attribute distributions. However, variations in attribute distribution may directly affect the filtering strategies and index construction efficiency of hybrid query algorithms.

@@

	(1)\textbf{ Systematic Classification and Overview.}
	We systematically classify 10 representative attribute filtering algorithms from multiple dimensions, including index organization, filtering strategy, Boolean logic support, and index construction methods. Additionally, we survey range filtering algorithms and provide an overview of widely used vector retrieval libraries (e.g., Faiss) and vector databases (e.g., Milvus, PASE, VBASE). These efforts provide a unified reference framework for future study.

	(2)\textbf{ Evaluation of Attribute Filtering Algorithms.}
	We conduct a systematic evaluation of 13 attribute filtering algorithms on 6 real-world datasets. By analyzing performance under varying numbers of attributes, we reveal the strengths and weaknesses of each method. We further examine their behavior under different attribute distributions and selectivity conditions to assess robustness and adaptability in complex query scenarios. Evaluation metrics include index construction time, index size, peak memory usage, QPS, and search accuracy.

	(3)\textbf{ Evaluation of Range Filtering Algorithms.}
	We benchmark 5 mainstream range filtering algorithms on 3 large-scale datasets, using varying query range settings in the experiments. The experimental results demonstrate the performance of these algorithms in terms of index construction efficiency, storage overhead, and query performance. Additionally, we conduct an in-depth analysis of how index organization impacts algorithm performance, providing valuable insights for algorithm design.

	(4)\textbf{ Recommendations and Challenges.}
	Based on the experimental results, we provide algorithm selection recommendations for common application scenarios and highlight key challenges in the field of hybrid queries. These challenges include limited Boolean logic support, the lack of multi-attribute range filtering capabilities, the high indexing costs of graph-based methods, and the sensitivity of algorithms to data distribution. Currently, few methods simultaneously support both attribute filtering and range filtering, pointing to potential directions for future research.

@@

	\subsection{Attribute Filtering Algorithms}

	We summarize 6 attribute filtering algorithms and provide an intuitive comparison in Table~\ref{tab:compair_1}.

	\noindent\textbf{\underline{NHQ \cite{NHQ}.}}  
	Traditional attribute filtering algorithms usually perform attribute constraints and ANN search separately. In contrast, NHQ is the first to implement simultaneous filtering, integrating both aspects into a unified framework. NHQ constructs an index based on a nearest neighbor graph and introduces a fusion distance that jointly captures vector similarity and attribute similarity. Leveraging this fusion distance, NHQ unifies vector similarity and attribute matching into a single comprehensive similarity measure and builds the graph accordingly. During the query process, NHQ efficiently prunes irrelevant edges via this composite index, enabling fast retrieval of results that satisfy both vector similarity and attribute constraints.

@@

	\subsection{Range Filtering Algorithms}

	We introduce 5 representative range filtering algorithms and present a comparative overview in Table~\ref{tab:range_algo}.

@@

	\noindent\textbf{\underline{WinFilter \cite{winFilter}.}}  
	Unlike SeRF, which focuses on graph compression, WinFilter proposes a structural partitioning framework called the $\beta$-Window Search Tree ($\beta$-WST). After the dataset is sorted by attribute values, it is partitioned into multiple intervals and organized into a tree structure. Each node in the tree corresponds to an attribute range and maintains a local ANN index (e.g., Vamana). For a range filtering query, WinFilter only searches within nodes overlapping the query range and merges partial results. With a tree height of $O(\log n)$, each query accesses at most $O(\log n)$ sub-indexes, resulting in significant speedups.

@@

	For attribute filtering tasks, we employ 6 real-world datasets widely adopted in existing literature: Msong~\cite{msong2011}, Audio~\cite{audio_unknown}, SIFT1M~\cite{sift2010}, GIST1M~\cite{sift2010}, GloVe~\cite{GloVe2015}, and Enron~\cite{enron2015}, covering a diverse range of domains. We independently generate attribute for each dataset.

	For range filtering queries, we evaluate 3 widely used real-world datasets: Deep~\cite{yandex_deep_dataset}, YT-Audio~\cite{youtube8m_dataset}, and WIT~\cite{wit_dataset}. Following prior work~\cite{DSG}, we use the data point ID within each dataset as its attribute. This ensures that the data is initially sorted.

@@

	\subsection{Range Filtering}
	Although the libraries and databases mentioned above can also implement range filtering, the results are poor, so we only compare 5 range filtering algorithms.
	\subsubsection{Time and Space overhead of index construction.}
	\begin{figure}[t]
		\centering
		\setlength{\abovecaptionskip}{0.3cm}
		\setlength{\belowcaptionskip}{0cm}

@@

	Similar to the evaluation of \textit{Attribute Filtering} algorithms, we also evaluate 3 key metrics mentioned above in \textit{Range Filtering} algorithms.

	\textit{\textbf{Index construction time.}}
	Figure~\ref{fig:rangeFilter_build_time} shows the single-threaded index construction time. Faiss builds the index fastest. Because Faiss obtains subsamples through sampling and then calculates the cluster centroid, the computational complexity is significantly reduced.

	WinFilter and ACORN-\(\gamma\) take the longest.
	WinFilter relies on a tree structure, requiring separate nearest-neighbor graphs per node. This causes redundant computation and high costs, especially for large datasets.
	ACORN-\(\gamma\) expands the neighbor list of each node to provide more candidate paths, which increases both computation and storage costs. It also prunes distant neighbors to save storage space; however, this introduces additional overhead and slows down the construction process.

	\textit{\textbf{Index size.}} 
	As shown in Figure~\ref{fig:rangeFilter_index_size_mb}, Among all algorithms, Faiss has the smallest index structure. Because IVF only adds centroid data and partition information to the vector dataset, the IVF index is only slightly larger than the vector dataset.
	On low-dimensional datasets (Deep, YT-Audio), PASE requires the most storage. Its implementation likely wastes space due to alignment padding and inefficient encoding (e.g., Base64).

	SeRF maintains a compact index for low-dimensional data. However, for high-dimensional datasets (WIT), SeRF stores more edges and dynamic range data, significantly increasing its size.

	\textit{\textbf{Peak memory usage.}}
	As Figure~\ref{fig:rangeFilter_memory_mb} shows, Faiss uses the least memory on low-dimensional datasets (Deep, YT-Audio). It stores minimal temporary data (e.g., vectors, centroids, distances) during index construction.

	DSG and WinFilter require significantly more memory than other methods.Because……

	For the high-dimensional dataset WIT, iRange achieves the lowest peak memory. It pre-builds graphs only for select intervals, skipping full-range index construction. This design keeps the index compact even in high dimensions.

	In contrast, Unify consumes the most memory because...
	\begin{figure*}[htbp]
		\centering
		%  \setlength{\abovecaptionskip}{0cm}
		%    \setlength{\belowcaptionskip}{-0.4cm}
		\includegraphics[width=\textwidth]{figures/exp/exp_8_2.pdf}
		\caption{RF-ANN search }
		\label{fig:exp_8_2}
	\end{figure*}

@@

%	 修改此处
	ACORN performs similarly on both large and small range queries, but its overall performance is worse than the other methods. Unlike other range filtering algorithms, ACORN is not designed with specialized data structures (such as the $\beta$-WST in WinFilter, the rectangle tree and dynamic segment graph in SeRF, or the HSIG in UNIFY) to achieve efficient range filtering. In fact, the graph structure used by ACORN for attribute filtering is the same as that used for range filtering.

@@

\subsection{Challenges}
\subsubsection{\textbf{Attribute Filtering}}
Most existing attribute filtering algorithms are graph-based, they typically outperform other methods in terms of query accuracy and efficiency. Filtered-DiskANN leverages disk-based storage, making it more suitable for extremely large-scale datasets even in memory-constrained environments. Moreover, the index construction phase of graph-based algorithms is computationally intensive. Therefore, exploring GPU-based acceleration for graph construction and vector computation during indexing is promising. It can significantly improve the performance of graph-based methods.

Furthermore, current attribute filtering algorithms offer limited support for complex filtering conditions. For example, Filtered-DiskANN  supports single-attribute filtering, and in multi-attribute scenarios, it only supports OR conditions between attributes. 
%The original implementations of 
NHQ imposes strict constraints on the number of base and query attributes and only support AND conditions. While UNG is relatively flexible—supporting both AND and OR logic—it still lacks support for arbitrary Boolean expressions (e.g., combinations of AND and OR). The lack of flexible Boolean filtering remains a key challenge, limiting the practicality of hybrid query methods in real-world systems with diverse and complex query requirements.

@@

Except for DSG, most range filtering algorithms require the dataset to be pre-sorted before building the index. Moreover, these methods do not support multi-attribute range queries. Enhancing these methods to support an arbitrary number of range filterable attributes remains an open research challenge.

@@

Lastly, our findings show that the distribution and cardinality of attributes, as well as dataset properties (e.g., dimensionality and LID), significantly impact the performance of attribute filtering methods. Enhancing the robustness of these algorithms under diverse data conditions is another key challenge that warrants further investigation.

@@

In this paper, we evaluate various hybrid query methods, including different algorithms, vector databases, and vector libraries. For attribute filtering algorithms, we design and conduct a series of experiments to comprehensively assess their overall performance. We then perform experiments on 6 real-world datasets to analyze the effectiveness of attribute filtering algorithms in depth. Additionally, we compare existing range filtering algorithms using 3 range query datasets. Finally, we provide a detailed analysis of the experimental results, summarize key findings, and highlight areas for improvement. Our study not only validates previous research but also offers insights for future work.

>>>>

	In recent years, hybrid query algorithms develop rapidly, giving rise to numerous attribute-filtering ~\cite{NHQ,diskann} and range-filtering algorithms \cite{serf,iRangeGraph}. In practical applications, the performance of hybrid query algorithms is influenced not only by unstructured data but also closely related to structured data. For attribute-filtering algorithms, factors such as the number of base attributes, the number of query attributes, attribute distribution \cite{UNG}, and attribute selectivity \cite{ACORN} significantly impact algorithm performance. As for range-filtering algorithms, the size of the query range and the characteristics of different datasets also play an important role in determining algorithm performance.

	Despite BigANN 2023 \cite{bigann2023} evaluates the performance of some algorithms, it presents several notable limitations:
	1) The evaluation covers a limited number of algorithms and fails to comprehensively include mainstream methods. For instance, attribute-filtering algorithms such as NHQ and UNG are not included in the evaluation. 2) It focuses only on single-attribute and dual-attribute query scenarios, neglecting the more complex demands of multi-attribute queries. In real-world applications, an object typically involves multiple attributes. For example, in e-commerce scenarios, users often specify multiple attributes (e.g., color, brand, price range) simultaneously when searching.
	3) It does not evaluate algorithm performance under different attribute distributions. However, variations in attribute distribution may directly affect the filtering strategies and index construction efficiency of hybrid query algorithms.

@@

	(1)\textbf{ Systematic Classification and Overview.}
	We systematically classify 10 representative attribute-filtering algorithms from multiple dimensions, including index organization, filtering strategy, Boolean logic support, and index construction methods. Additionally, we survey range-filtering algorithms and provide an overview of widely used vector retrieval libraries (e.g., Faiss) and vector databases (e.g., Milvus, PASE, VBASE). These efforts provide a unified reference framework for future study.

	(2)\textbf{ Evaluation of Attribute-Filtering Algorithms.}
	We conduct a systematic evaluation of 13 attribute filtering algorithms on 6 real-world datasets. By analyzing performance under varying numbers of attributes, we reveal the strengths and weaknesses of each method. We further examine their behavior under different attribute distributions and selectivity conditions to assess robustness and adaptability in complex query scenarios. Evaluation metrics include index construction time, index size, peak memory usage, QPS, and search accuracy.

	(3)\textbf{ Evaluation of Range-Filtering Algorithms.}
	We benchmark 5 mainstream range-filtering algorithms on 3 large-scale datasets, using varying query range settings in the experiments. The experimental results demonstrate the performance of these algorithms in terms of index construction efficiency, storage overhead, and query performance. Additionally, we conduct an in-depth analysis of how index organization impacts algorithm performance, providing valuable insights for algorithm design.

	(4)\textbf{ Recommendations and Challenges.}
	Based on the experimental results, we provide algorithm selection recommendations for common application scenarios and highlight key challenges in the field of hybrid queries. These challenges include limited Boolean logic support, the lack of multi-attribute range-filtering capabilities, the high indexing costs of graph-based methods, and the sensitivity of algorithms to data distribution. Currently, few methods simultaneously support both attribute filtering and range filtering, pointing to potential directions for future research.

@@

	\subsection{Attribute-Filtering Algorithms}

	We summarize 6 attribute-filtering algorithms and provide an intuitive comparison in Table~\ref{tab:compair_1}.

	\noindent\textbf{\underline{NHQ \cite{NHQ}.}}  
	Traditional attribute-filtering algorithms usually perform attribute constraints and ANN search separately. In contrast, NHQ is the first to implement simultaneous filtering, integrating both aspects into a unified framework. NHQ constructs an index based on a nearest neighbor graph and introduces a fusion distance that jointly captures vector similarity and attribute similarity. Leveraging this fusion distance, NHQ unifies vector similarity and attribute matching into a single comprehensive similarity measure and builds the graph accordingly. During the query process, NHQ efficiently prunes irrelevant edges via this composite index, enabling fast retrieval of results that satisfy both vector similarity and attribute constraints.

@@

	\subsection{Range-Filtering Algorithms}

	We introduce 5 representative range-filtering algorithms and present a comparative overview in Table~\ref{tab:range_algo}.

@@

	\noindent\textbf{\underline{WinFilter \cite{winFilter}.}}  
	Unlike SeRF, which focuses on graph compression, WinFilter proposes a structural partitioning framework called the $\beta$-Window Search Tree ($\beta$-WST). After the dataset is sorted by attribute values, it is partitioned into multiple intervals and organized into a tree structure. Each node in the tree corresponds to an attribute range and maintains a local ANN index (e.g., Vamana). For a range-filtering query, WinFilter only searches within nodes overlapping the query range and merges partial results. With a tree height of $O(\log n)$, each query accesses at most $O(\log n)$ sub-indexes, resulting in significant speedups.

@@

	For attribute-filtering tasks, we employ 6 real-world datasets widely adopted in existing literature: Msong~\cite{msong2011}, Audio~\cite{audio_unknown}, SIFT1M~\cite{sift2010}, GIST1M~\cite{sift2010}, GloVe~\cite{GloVe2015}, and Enron~\cite{enron2015}, covering a diverse range of domains. We independently generate attribute for each dataset.

	For range-filtering queries, we evaluate 3 widely used real-world datasets: Deep~\cite{yandex_deep_dataset}, YT-Audio~\cite{youtube8m_dataset}, and WIT~\cite{wit_dataset}. Following prior work~\cite{DSG}, we use the data point ID within each dataset as its attribute. This ensures that the data is initially sorted.

@@

	\subsection{Range Filtering}

	\subsubsection{Time and Space overhead of index construction.}
	%4.2.1

	\begin{figure}[t]
		\centering
		\setlength{\abovecaptionskip}{0.3cm}
		\setlength{\belowcaptionskip}{0cm}

@@

	Similar to the evaluation of \textit{Attribute-Filtering} algorithms, we also evaluate 3 key metrics mentioned above in \textit{Range-Filtering} algorithms.
	%serf: It incrementally builds the index in attribute order. It requires only a single pass and adds edges within local graphs. The structure is simple, with no redundant operations.

	\textit{\textbf{Index construction time.}}
	Figure~\ref{fig:rangeFilter_build_time} shows the single-threaded index construction time. Faiss builds the index fastest.

	WinFilter and ACORN-\(\gamma\) take the longest.
	WinFilter relies on a tree structure, requiring separate nearest-neighbor graphs per node. This causes redundant computation and high costs, especially for large datasets.

	ACORN-\(\gamma\) expands each node’s neighbor list to offer more candidate paths, increasing computation and storage. It also prunes distant neighbors to reduce storage, but this adds extra overhead, slowing construction.

	\textit{\textbf{Index size.}} 
	As shown in Figure~\ref{fig:rangeFilter_index_size_mb}, Among all algorithms, Faiss has the smallest index structure because. On low-dimensional datasets (deep, yt-audio), PASE has the largest index size because its implementation may have wasteful alignment and padding of data structures, as well as the use of storage methods such as Base64 encoding, which significantly increase storage overhead.
	The index size of SeRF on low-dimensional datasets is not large, but on high-dimensional datasets (wit), SeRF is forced to store more edges and dynamic range information, so the index increases significantly

	\textit{\textbf{Peak memory usage.}}
	As shown in Figure~\ref{fig:rangeFilter_memory_mb}, For low-dimensional datasets (deep, yt-audio), Faiss takes up the lowest memory due to: .

	On low-dimensional datasets, DSG and WinFilter exhibit higher peak memory usage compared to other algorithms. This suggests that their index structures suffer from severe structural redundancy and are difficult to compress in low-dimensional scenarios.
	%因为低维空间中数据点之间的欧氏距离普遍较小，点与点之间更易互为近邻，从而导致图索引中产生大量冗余的边连接。

	For high-dimensional datasets WIT, iRange achieves the lowest peak memory usage. It pre-builds graphs only for a limited number of intervals, avoiding explicit index construction for all possible query ranges. Its index structure remains stable in high dimensions without significant graph growth. This effectively controls index size, making iRange more advantageous for high-dimensional datasets.
	In contrast, Unify exhibits the highest memory consumption, indicating that its index construction process is highly sensitive to the dimensionality of the data.

	\begin{figure*}[htbp]
		\centering
		%  \setlength{\abovecaptionskip}{0cm}
		%    \setlength{\belowcaptionskip}{-0.4cm}
		\includegraphics[width=\textwidth]{figures/exp/exp_8_2.pdf}
		\caption{RF-ANN search }
		\label{fig:exp_8_2}
	\end{figure*}

@@

%	 修改此处
	ACORN performs similarly on both large and small range queries, but its overall performance is worse than the other methods. Unlike other range-filtering algorithms, ACORN is not designed with specialized data structures  to achieve efficient range filtering. In fact, the graph structure used by ACORN for attribute filtering is the same as that used for range filtering.

@@

\subsection{Challenges}
\subsubsection{\textbf{Attribute Filtering}}
Most existing attribute-filtering algorithms are graph-based, they typically outperform other methods in terms of query accuracy and efficiency. Filtered-DiskANN leverages disk-based storage, making it more suitable for extremely large-scale datasets even in memory-constrained environments. Moreover, the index construction phase of graph-based algorithms is computationally intensive. Therefore, exploring GPU-based acceleration for graph construction and vector computation during indexing is promising. It can significantly improve the performance of graph-based methods.

Furthermore, current attribute-filtering algorithms offer limited support for complex filtering conditions. For example, Filtered-DiskANN  supports single-attribute filtering, and in multi-attribute scenarios, it only supports OR conditions between attributes. 
%The original implementations of 
NHQ imposes strict constraints on the number of base and query attributes and only support AND conditions. While UNG is relatively flexible—supporting both AND and OR logic—it still lacks support for arbitrary Boolean expressions (e.g., combinations of AND and OR). The lack of flexible Boolean filtering remains a key challenge, limiting the practicality of hybrid query methods in real-world systems with diverse and complex query requirements.

@@

Except for DSG, most range-filtering algorithms require the dataset to be pre-sorted before building the index. Moreover, these methods do not support multi-attribute range queries. Enhancing these methods to support an arbitrary number of range-filterable attributes remains an open research challenge.

@@

Lastly, our findings show that the distribution and cardinality of attributes, as well as dataset properties (e.g., dimensionality and LID), significantly impact the performance of attribute-filtering methods. Enhancing the robustness of these algorithms under diverse data conditions is another key challenge that warrants further investigation.

@@

In this paper, we evaluate various hybrid query methods, including different algorithms, vector databases, and vector libraries. For attribute-filtering algorithms, we design and conduct a series of experiments to comprehensively assess their overall performance. We then perform experiments on 6 real-world datasets to analyze the effectiveness of attribute-filtering algorithms in depth. Additionally, we compare existing range-filtering algorithms using 3 range query datasets. Finally, we provide a detailed analysis of the experimental results, summarize key findings, and highlight areas for improvement. Our study not only validates previous research but also offers insights for future work.
